---
title: "Introduction to data"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    css: ./lab.css
---

Some define Statistics as the field that focuses on turning information into
knowledge.  The first step in that process is to summarize and describe the raw
information - the data.  In this lab, you will gain insight into public health
by generating simple graphical and numerical summaries of a data set collected 
by the Centers for Disease Control and Prevention (CDC).  As this is a large 
data set, along the way you'll also learn the indispensable skills of data 
processing and subsetting.


## Getting started

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone 
survey of 350,000 people in the United States. As its name implies, the BRFSS 
is designed to identify risk factors in the adult population and report 
emerging health trends. For example, respondents are asked about their diet and 
weekly physical activity, their HIV/AIDS status, possible tobacco use, and even
their level of healthcare coverage. The BRFSS Web site 
([http://www.cdc.gov/brfss](http://www.cdc.gov/brfss)) contains a complete 
description of the survey, including the research questions that motivate the 
study and many interesting results derived from the data.

We will focus on a random sample of 20,000 people from the BRFSS survey 
conducted in 2000. While there are over 200  variables in this data set, we will
work with a small subset.

We begin by loading the data set of 20,000 observations into the R workspace. 
After launching RStudio, enter the following command.

```{r load-data, eval=TRUE}
source("more/cdc.R")
```

The data set `cdc` that shows up in your workspace is a *data matrix*, with each
row representing a *case* and each column representing a *variable*.  R calls 
this data format a *data frame*, which is a term that will be used throughout 
the labs.

To view the names of the variables, type the command

```{r names, eval=TRUE}
names(cdc)
```

This returns the names `genhlth`, `exerany`, `hlthplan`, `smoke100`, `height`, 
`weight`, `wtdesire`, `age`, and `gender`. Each one of these variables 
corresponds to a question that was asked in the survey.  For example, for 
`genhlth`, respondents were asked to evaluate their general health, responding
either excellent, very good, good, fair or poor. The `exerany` variable 
indicates whether the respondent exercised in the past month (1) or did not (0).
Likewise, `hlthplan` indicates whether the respondent had some form of health 
coverage (1) or did not (0). The `smoke100` variable indicates whether the 
respondent had smoked at least 100 cigarettes in her lifetime. The other 
variables record the respondent's `height` in inches, `weight` in pounds as well
as their desired weight, `wtdesire`, `age` in years, and `gender`.

1.  How many cases are there in this data set?  How many variables?  For each 
    variable, identify its data type (e.g. categorical, discrete).
```{r A1, eval=TRUE}
dim(cdc)
sapply(cdc, class)
```
The datatypes for each variable are:
genhealth: categorical, ordinal;
exerany: categorical, nominal;
hlthplan: categorical, nominal;
smoke100: categorical, nominal;
height: numerical, discrete;
weight: numerical, discrete;
wtdesire: numerical, discrete;
age: numerical, discrete;
gender: categorical, nominal

## Summaries and tables

2.  Create a numerical summary for `height` and `age`, and compute the 
    interquartile range for each. Compute the relative frequency distribution for
    `gender` and `exerany`. How many males are in the sample? What proportion of
    the sample reports being in excellent health?
    
```{r A2, eval=TRUE}
summary(cdc$height)
summary(cdc$age)
IQR(cdc$height)
IQR(cdc$age)
table(cdc$gender)
prop.table(table(cdc$gender))
table(cdc$exerany)
prop.table(table(cdc$exerany))
prop.table(table(cdc$genhlth))
```


```{r mosaic-smoke-gender, eval=TRUE}
mosaicplot(table(cdc$gender,cdc$smoke100))
```

We could have accomplished this in two steps by saving the table in one line and
applying `mosaicplot` in the next (see the table/barplot example above).

3.  What does the mosaic plot reveal about smoking habits and gender?

A3: The mosaic plot reveals that more males have smoked 100 cigarettes in their lifetime than females.

## Interlude: How R thinks about data

4.  Create a new object called `under23_and_smoke` that contains all observations 
    of respondents under the age of 23 that have smoked 100 cigarettes in their 
    lifetime. Write the command you used to create the new object as the answer
    to this exercise.
    
```{r}
under23_and_smoke <- subset(cdc, age < 23 & smoke100 == 1)
```

## Quantitative data


```{r boxplot-bmi, eval=TRUE}
bmi <- (cdc$weight / cdc$height^2) * 703
boxplot(bmi ~ cdc$genhlth)
```

Notice that the first line above is just some arithmetic, but it's applied to 
all 20,000 numbers in the `cdc` data set. That is, for each of the 20,000 
participants, we take their weight, divide by their height-squared and then 
multiply by 703. The result is 20,000 BMI values, one for each respondent. This 
is one reason why we like R: it lets us perform computations like this using 
very simple expressions.

5.  What does this box plot show? Pick another categorical variable from the 
    data set and see how it relates to BMI. List the variable you chose, why you
    might think it would have a relationship to BMI,  and indicate what the 
    figure seems to suggest. 
    
5A: The above box plot of BMI vs general health shows a trend that as BMI increases the persons general health is more likely to be worse. Below I have created a box plot that displays BMI vs exerany variable. I chose this because exercise is a proven method to reduce a persons weight and if someone reported that they have exercised in the past month they might have a lower BMI than someone who does not exercise. The below figure seems to confirm my hypothesis, although not definitively, because the median and box show a lower BMI for respondants that have exercised.

```{R 4A, eval=TRUE}
boxplot(bmi ~ cdc$exerany)
```


* * *

## On Your Own

-   Make a scatterplot of weight versus desired weight. Describe the 
    relationship between these two variables.
    
```{r scatter plot, eval=TRUE}
plot(cdc$wtdesire~cdc$weight)
```
The above scatter plot shows that the majority of respondants desire a weight between 100 and 250 pounds and desire a weight less than their current weight. We also can see there are a couple outliers where the respondants current weight is quite high or that their desired weight is extremely high.

-   Let's consider a new variable: the difference between desired weight 
    (`wtdesire`) and current weight (`weight`). Create this new variable by 
    subtracting the two columns in the data frame and assigning them to a new 
    object called `wdiff`.
    
```{r wdiff, eval=TRUE}
wdiff <- cdc$weight - cdc$wtdesire
```

-   What type of data is `wdiff`? If an observation `wdiff` is 0, what does 
    this mean about the person's weight and desired weight. What if `wdiff` is 
    positive or negative?
    
wdiff is a discrete numerical data type where a value of 0 indicates the respondant was currently at their desired weight. If wdiff is positive that indicates that the respondant would like to lose weight; if it is negative it indicates that the respondant would like to gain weight

-   Describe the distribution of `wdiff` in terms of its center, shape, and 
    spread, including any plots you use. What does this tell us about how people 
    feel about their current weight?
```{r wdiff plots, eval=TRUE}
summary(wdiff)
boxplot(wdiff)
hist(wdiff, breaks = 30)
```
Looking at the summary, boxplot and histogram we can see that there are outliers that are far away from the median of 10. The boxplot is fairly hard to see as the spread is so large and the majority of respondants within the interquartile range are near 0. Looking at a histogram with 30 bins, it is a bit easier to see that most respondants are satisfied with their weight but more would like to lose a little weight than gain weight.

-   Using numerical summaries and a side-by-side box plot, determine if men tend
    to view their weight differently than women.
    
```{r, eval=TRUE}
summary(wdiff)
boxplot(wdiff ~ cdc$gender)
```

-   Now it's time to get creative. Find the mean and standard deviation of 
    `weight` and determine what proportion of the weights are within one 
    standard deviation of the mean.
    
```{r}
sd(cdc$weight)
mean(cdc$weight)
sub <- subset(cdc, weight < (mean(cdc$weight) + sd(cdc$weight)) & weight > (mean(cdc$weight) - sd(cdc$weight)))
nrow(sub)/nrow(cdc)
```

<div id="license">
This is a product of OpenIntro that is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel 
from a lab written by Mark Hansen of UCLA Statistics.
</div>
